\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology

### 2022-06-10 jsdse/reviews/response_2022_06_09.Rmd

---
output: rmarkdown::pdf_document
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
geometry: margin=1.2in
fontsize: 12pt
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

January 10, 2022

University of California, Merced

&nbsp;  &nbsp;

Dear Dr. Horton, associate editor, and reviewers,

&nbsp;

We appreciate the helpful and thoughtful comments made to our revised manuscript. We are excited to learn that the manuscript is now suitable for publication, subject to minor revisions.

Please receive a newly revised version of our manuscript, formerly titled *“Promoting analysis reproducibility with accessibility: An example in evolutionary biology”*, and now titled **“Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology”**.

We have incorporated into the revised manuscript all comments and suggestions from you, associate editor and 2 reviewers. We believe that all of them have greatly improved the manuscript. We hope that you find it ready for publication in the special issue about teaching reproducibility and responsible workflow.

We address individual comments line by line below.

&nbsp;  &nbsp;

Please receive our respectful regards,

&nbsp;  &nbsp;

The authors

\pagebreak

\singlespacing

*Comments from the Editors and Reviewers:*

*Editor*
*Comments to the Author*

*In addition to the comments from the reviewers and associate editor, I had the following additional points:*

*1. The abstract (in the metadata associated with the paper) is missing spacing in many places (e.g., "reproducibilityare". This is likely due to cut and paste issues. Can you please review and fix as this will promulgate through the production process.*

Thanks. We reviewed and fixed missing spaces across (hopefully) the entire manuscript.

&nbsp;  &nbsp;

*2. It might be appropriate to consider citing this in-process work (https://arxiv.org/abs/2202.09504)*

We agree. We have cited this work...

&nbsp;  &nbsp;

*3. page 2, end of third paragraph: would it be useful to cite the NASEM 2018 (Data Science for Undergraduates) report, which includes reproducibility and workflow as important components of "data acumen"? In addition, would citation of the NASEM 2019 Reproducibility report be appropriate?*

Yes, after pondering your question, we concluded that the NASEM 2018 report should also be cited in this section. The NASEM 2019, I am not completely sure yet...

&nbsp;  &nbsp;

*4. Note style for journal (Taylor & Francis) is to capitalize article and book titles*

Thanks. We went over the entire bib file and capitalized article and book titles. Hopefully they are all appropriately styled now.

&nbsp;  &nbsp;

*********************

*Associate Editor*
*Comments to the Author:*
*Thank you for reviewing the manuscript according to reviewers feedback. One issue which is pointed by one reviewer is mentioning package version - which is your first section in the online resource you discuss, so it will be easy to incorporate.*

Thank you! We addressed that on section...

&nbsp;  &nbsp;

*In the Conclusions when you provide examples of some institutions adopting reproducibility as a main teaching principle, you could also mention University of Glasgow School of Psychology and Neuroscience - https://psyteachr.github.io/*

Thanks for pointing us out to this resource. We thought it was appropriate, so we added psyteachR to the examples.

&nbsp;  &nbsp;

*********************

*Reviewer: 1*

*Comments to the Author*
*The authors have addressed all the issues I had raised about the initial submission.*

We really appreciate all comments and suggestions made on the initial review!

&nbsp;  &nbsp;

*********************

*Reviewer: 2*

*Comments to the Author*
*A significant barrier not discussed here is the deprecation of functions in open software packages like R and Python. It often requires considerable programming savvy to repair code that has fallen victim to deprecation.*

Thank you, the point that you raise is very important. Associate Editor correctly points out that this can be addressed by providing information on package version used. In general, old package versions remain accessible, and persist in time, so they can be used to reproduce an analysis, even long time after a function had been deprecated. 
The teaching materials that we use as example include a section that addresses documenting package versions used for an analysis.

We elaborate on this barrier on section...

&nbsp;  &nbsp;

*Line 38: considerably* - Fixed

*Line 54: “…as a model…”* - Fixed

*Line 30: Define “live programming”* - Added a sentence defining it.

*Line 37: This is called Carpentries here but Software Carpentries in Figure 3’s caption. Are these the same thing?*  - Software Carpentries is a component of the Carpentries, we clarified this in the Figure caption.

*Line 53 and elsewhere: “learner’s” and “user’s”.* - Fixed

*Line 12: Warner et al. 2021 do not make this claim in their paper.* - You are right, we meant...

*********************

### 2022-06-10 jsdse/reviews/response_2022_06_09.Rmd

We agree. We have cited this work...

&nbsp;  &nbsp;

*3. page 2, end of third paragraph: would it be useful to cite the NASEM 2018 (Data Science for Undergraduates) report, which includes reproducibility and workflow as important components of "data acumen"? In addition, would citation of the NASEM 2019 Reproducibility report be appropriate?*

Yes, after pondering your question, we concluded that the NASEM 2018 report should also be cited in this section.

The NASEM 2019 report is very interesting and addresses many important points relevant for this work. We cite it at the end of first paragraph, Page 5: "As computers continue to play a larger role in most scientific disciplines (Piccolo & Frampton 2016), higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers (NASEM, 2019)"

@@

*Line 30: Define “live programming”* - Added a sentence defining it.

*Line 37: This is called Carpentries here but Software Carpentries in Figure 3’s caption. Are these the same thing?*  - Software Carpentries is a component of the Carpentries, we clarified this in the Figure caption.

*Line 53 and elsewhere: “learner’s” and “user’s”.* - Fixed

*Line 12: Warner et al. 2021 do not make this claim in their paper.* - You are right, we changed the sentence accordingly to: "[The principles described here] can also help close the academic gap that is generated by uneven access to computational resources across students belonging to different groups (KewalRamani et al. 2018), where the most affected learners usually belong to underrepresented minorities and rural areas (Warner et al. 2021)."

### 2022-06-10 jsdse/Manuscript.tex
response to reviews mostly done!

In this paper, we argue that research products that are digitally available are often unapproachable in practice, because they are not friendly for learners and interested parties with different levels of experience in the field.
Research products that are unapproachable counteract availability, and hinder reproducibility short and long term.
To support long term adoption of reproducible practices in the natural sciences, research workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think}.

To elaborate on our thesis, we designed a case study within the research field of phylogenetics, a discipline within evolutionary biology.
We use our case study to identify barriers that have made research workflows largely unapproachable to a general audience in the natural sciences.
Then, we propose some principles for researchers to address these barriers, and create research workflows that are reproducible by a larger audience.
% ** describe/introduce reproducibility curriculum here
The principles proposed here can be generalized and integrated into the undergraduate and graduate school STEM curriculum, either for courses specialized in reproducibility or within other subject areas, as a necessary component of successful and impactful science.

@@

However, while learners in the natural sciences have been engaging independently with R and Python programming languages, computer programming is not traditionally a core skill formally taught to biologists and naturalists \citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools}, higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers.

@@

The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}.
If a researcher were to use the proposed analysis workflow in a publication, they would typically describe it in the methodology section as ``The canid summary phylogeny was obtained using functions from X package, details are available as supplementary files".
This is usual practice, mainly because journals do not have space to publish all code used for an analysis in the methods section. Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
Without the primary data and code that was used to perform an analysis it is impossible to reproduce said analysis \citep{miyakawa2020no}. When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
For example, some questions that are often unaddressed in research workflows are:
\textit{what software do I need to open the code files?}
\textit{Can I run the code with the same software I used to open the files, or do I need a different one?}
\textit{Do I need additional software that the analysis depends on?}
\textit{What does the code even mean?}

@@

A third important aspect of software documentation are examples that demonstrate usage of individual functions \citep{karimzadeh2018top}.
Examples presented in software documentation are usually worked to perfection, as they are intended to showcase the ideal or minimal case in which a function works well.
Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
We identify this as \textit{barrier 3.-- Content-focused examples}.
Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
actively explore a task with a positive mindset \citep{frese1995error}.
Providing examples that showcase potential errors supports user's performance \citep{steele2014error}, and can greatly improve learner's ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.

@@

The principles to create tutorials described here facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
It will also help close the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
Late access to computational resources and training can occur due to lack of economic resources, often occurring in households from underrepresented communities and minorities \citep{google2016diversity, warner2021quantifying}.
It can also be due to gender-biased parental and community pressures, in which male identifying individuals are more often encouraged to perform activities related to computers, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
These principles can be used to improve not only reproducibility practices, but also software adoption in the natural sciences.

>>>>

In this paper, we argue that research products that are digitally available are often unapproachable in practice, because they are not friendly for learners and interested parties with different levels of experience in the field.
Research products that are unapproachable counteract availability, and hinder reproducibility short and long term.
To support long term adoption of reproducible practices in the natural sciences, research workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think, nasem2018data}.

To elaborate on our thesis, we designed a case study within the research field of phylogenetics, a discipline within evolutionary biology.
We use our case study to identify barriers that have made research workflows largely unapproachable to a general audience in the natural sciences.
Then, we propose some principles for researchers to address these barriers, and create research workflows that are reproducible by a larger audience.
% ** describe/introduce reproducibility curriculum here:
The principles proposed here can be generalized and integrated into the undergraduate and graduate school STEM curriculum, either for courses specialized in reproducibility or within other subject areas, as a necessary component of successful and impactful science.

@@

However, while learners in the natural sciences have been engaging independently with R and Python programming languages, computer programming is not traditionally a core skill formally taught to biologists and naturalists \citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools}, higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers \citep{nasem2019reproducibility}.

@@

The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}.
If a researcher were to use the proposed analysis workflow in a publication, they would typically describe it in the methodology section as ``The canid summary phylogeny was obtained using functions from X package, details are available as supplementary files".
This is usual practice, mainly because journals do not have space to publish all code used for an analysis in the methods section. Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
Without the primary data and code that was used to perform an analysis, it is impossible to reproduce said analysis \citep{miyakawa2020no}. When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
For example, some questions that are often unaddressed in research workflows are:
\textit{what software do I need to open the code files?}
\textit{Can I run the code with the same software I used to open the files, or do I need a different one?}
\textit{Do I need additional software that the analysis depends on?}
\textit{What does the code even mean?}

@@

A third important aspect of software documentation are examples that demonstrate usage of individual functions \citep{karimzadeh2018top}.
Examples presented in software documentation are usually worked to perfection, as they are intended to showcase the ideal or minimal case in which a function works well.
Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
We identify this as \textit{barrier 3.-- Content-focused examples}.
Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
actively explore a task with a positive mindset \citep{frese1995error}.
Providing examples that showcase potential errors, supports user's performance \citep{steele2014error}, and can greatly improve learner's ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.

@@

The principles described here are relevant to facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
They can also help close the academic gap that is generated by uneven access to computational resources across students belonging to different groups \citep{kewalramani2018student}, where the most affected learners usually belong to underrepresented minorities and rural areas \citep{warner2021quantifying}.
Differences in access can also be due to gender-biased parental and community pressures, in which male identifying individuals are more likely to be encouraged to perform activities related to computers \citep{google2016diversity}, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
These principles can be used to improve not only reproducibility practices, but also ease adoption of computational resources in the natural sciences.
